Capstone Movielens Report

### 2025-12-22 reports/capstone-movielens.site/02-3.OMR-model.Rmd

We know that the estimate that minimizes the RMSE is the least squares estimate of $\mu$ and, in this case, is the average of all ratings:

```{r pressure, echo=TRUE}
mu <- mean(edx$rating)
print(mu)
```

If we predict all unknown ratings with $\hat{\mu}$, we can obtain RMSE as follows:: 
```{r}
mu.MSEs <- naive_model_MSEs(mu)
data.frame(fold_No = 1:5, MSE = mu.MSEs) |>
  data.plot(title = "MSE resuls of the 5-fold CV method applied to the Overall Mean Rating Model",
              xname = "fold_No", 
              yname = "MSE")

@@

  - [naive_model_MSEs](#func.naive_model_MSEs) (described in section [OMR Model: Helper Functions]) to compute *Mean Squared Errors* with the use of *K-Fold Cross-Validation*, where the $K$ is the length of the [`edx_CV` Object] (described in detail in [Appendix B: Models Training Datasets]) to which the *K-Fold Cross-Validation* is applied (in *this Project* we use $K = 5$);

  - [data.plot](#func.data.plot) (described in section [Data Visualization Functions]) for the _Mean Squared Errors_ data visualization.

@@

If we plug in any other number, we will get a higher *RMSE*. Let's prove that by the following small investigation:
```{r }
  deviation <- seq(0, 6, 0.1) - 3

@@

Let's make a quick investigation of the `deviation.RMSE` result we have just got: 
```{r}
data.frame(delta = deviation, 
           delta.RMSE = deviation.RMSE) |> 
data.plot(title = TeX(r'[RMSE as a function of deviation ($\delta$) from the Overall Mean Rating ($\hat{mu}$)]'),
              xname = "delta", 
              yname = "delta.RMSE", 
              xlabel = TeX(r'[$\delta$]'), 
              ylabel = "RMSE")

@@

```{r eval=FALSE}
RMSEs.ResultTibble.OMR <- RMSEs.ResultTibble |> 
  RMSEs.AddRow("Overall Mean Rating Model", mu.RMSE)
```
```{r}
RMSE_kable(RMSEs.ResultTibble.OMR)
```

>>>>

We know that the estimate that minimizes the RMSE is the least squares estimate of $\mu$ and, in this case, is the average of all ratings. We calculate this using the folowing [line of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L106):
```{{r main-script-l106, echo=TRUE, }
mu <- mean(edx$rating)
mu
```

If we predict all unknown ratings with $\hat{\mu}$, we can obtain *RMSE* using the following [code snippet](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L108): 
```{r main-script-l08 }
mu.MSEs <- naive_model_MSEs(mu)
data.frame(fold_No = 1:5, MSE = mu.MSEs) |>
  data.plot(title = "MSE resuls of the 5-fold CV method applied to the Overall Mean Rating Model",
              xname = "fold_No", 
              yname = "MSE")

@@

  - [naive_model_MSEs](#func.naive_model_MSEs) (described in section [OMR Model: Helper Functions]) to compute *Mean Squared Errors* (hereafter *MSE* for short) with the use of *K-Fold Cross-Validation*, where the $K$ is the length of the [`edx_CV` Object] (described in detail in [Appendix B: Models Training Datasets]) to which the *K-Fold Cross-Validation* is applied (in *this Project* we use $K = 5$);

  - [data.plot](#func.data.plot) (described in section [Data Visualization Functions]) for the *MSE* data visualization.

@@

If we plug in any other number, we will get a higher *RMSE*. Let's prove this with a small experiment using the following [code snippet](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L155):
```{r main-script-l155 }
  deviation <- seq(0, 6, 0.1) - 3

@@

Let's quickly analyze the resulting `deviation.RMSE` vector we have just got. The following [data visualization](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L189) code will help us with this:
```{r main-script-l189}
data.frame(delta = deviation, 
           delta.RMSE = deviation.RMSE) |> 
data.plot(title = TeX(r'[RMSE as a function of deviation ($\delta$) from the Overall Mean Rating ($\hat{mu}$)]'),
              xname = "delta", 
              yname = "delta.RMSE", 
              xlabel = TeX(r'[$\delta$]'), 
              ylabel = "RMSE")

@@

Finally, we add the *RMSE* value calculated for the *OMR Model* to our *Result Table* and print the table using the following [piece of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L200):
```{r main-script-l200, eval=FALSE}
RMSEs.ResultTibble.OMR <- RMSEs.ResultTibble |> 
  RMSEs.AddRow("Overall Mean Rating Model", mu.RMSE)
```
```{r}
RMSE_kable(RMSEs.ResultTibble.OMR)
```

### 2025-12-22 reports/capstone-movielens.site/02.4.UE-model.Rmd

These considerations alows us to compute a _User Mean Ratings_ the following way:
```{r eval=FALSE}
put_log("Computing Average Ratings per User (User Mean Ratings)...")
user.mean_ratings <- rowMeans(edx.mx, na.rm = TRUE)
user_ratings.n <- rowSums(!is.na(edx.mx))

@@

And then we compute a _User Effect_ this way:
```{r eval=FALSE}
put_log("Computing User Effect per users ...")
edx.user_effect <- edx.user_mean_ratings |>
  mutate(userId = as.integer(userId),
         a = mean_rating - mu)

@@

Now, we are ready to compute the _Mean Squared Errors_ for samples used in _K-Fold Cross-Validation_ (additionally using the [clamp](#func.clamp) helper function (described in the [Utility Functions] section of [Appendix A](#appndx_a)).

@@

The code that performs this operation is shown below:
```{r eval=FALSE}
put_log("Computing the RMSE taking into account user effects...")
edx.user_effect.MSEs <- sapply(edx_CV, function(cv_fold_dat){
  cv_fold_dat$validation_set |>
    left_join(edx.user_effect, by = "userId") |>
    mutate(resid = rating - clamp(mu + a)) |> 
    pull(resid) |> mse()
})
```
```{r}
data.frame(fold_No = 1:5, MSE = edx.user_effect.MSEs) |>
  data.plot(title = "MSE resuls of the 5-fold CV method applied to the User Effect Model",
            xname = "fold_No", 
            yname = "MSE")

@@

::: {.noteblock data-latex=""}
For the _Mean Squared Errors_ data visualization we use the custom data visualization function [data.plot](#func.data.plot) described in the [Data Visualization Functions](#appndx_a.data_helper.functions) section of [Appendix A](#appndx_a).
:::

Finally, we calculate the `RMSE` as the _square root_ of the average of the _Mean Squared Errors_ we obtained through the _5-Fold Cross-Validation_ above:   
```{r, eval=FALSE}
edx.user_effect.RMSE <- sqrt(mean(edx.user_effect.MSEs))

RMSEs.ResultTibble.UE <- RMSEs.ResultTibble.OMR |> 
  RMSEs.AddRow("User Effect Model", edx.user_effect.RMSE)
```
```{r}
RMSE_kable(RMSEs.ResultTibble.UE)
```

>>>>

These considerations allow us to compute a _User Mean Ratings_ using the following [piece code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L231):
```{r eval=FALSE}
put_log("Computing Average Ratings per User (User Mean Ratings)...")
user.mean_ratings <- rowMeans(edx.mx, na.rm = TRUE)
user_ratings.n <- rowSums(!is.na(edx.mx))

@@

And then we compute a _User Effect_ using the following [piece of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L290):
```{r eval=FALSE}
put_log("Computing User Effect per users ...")
edx.user_effect <- edx.user_mean_ratings |>
  mutate(userId = as.integer(userId),
         a = mean_rating - mu)

@@

Now, we are ready to compute the *Mean Squared Errors (MSE)* for samples used in _K-Fold Cross-Validation_ (additionally using the [clamp](#func.clamp) helper function (described in the [Utility Functions] section of [Appendix A](#appndx_a)).

@@

The [code snippet](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L345) that performs this operation is shown below:
```{r eval=FALSE}
put_log("Computing the RMSE taking into account user effects...")
edx.user_effect.MSEs <- sapply(edx_CV, function(cv_fold_dat){
  cv_fold_dat$validation_set |>
    left_join(edx.user_effect, by = "userId") |>
    mutate(resid = rating - clamp(mu + a)) |> 
    pull(resid) |> mse()
})
```
```{r}
data.frame(fold_No = 1:5, MSE = edx.user_effect.MSEs) |>
  data.plot(title = "MSE resuls of the 5-fold CV method applied to the User Effect Model",
            xname = "fold_No", 
            yname = "MSE")

@@

::: {.noteblock data-latex=""}
For the *MSE* data visualization we use the custom data visualization function [data.plot](#func.data.plot) described in the [Data Visualization Functions](#appndx_a.data_helper.functions) section of [Appendix A](#appndx_a).
:::

We can now calculate the *Root Mean Squared Error (RMSE)* for the current model as a *square root* of the average of the *MSE* values we obtained above, using the following [line of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L364):
```{r, eval=FALSE}
edx.user_effect.RMSE <- sqrt(mean(edx.user_effect.MSEs))
```

Finally, we add the *RMSE* value calculated for the *UE Model* to our *Result Table* and print the table using the following [piece of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L372):
```{r, eval=FALSE}
RMSEs.ResultTibble.UE <- RMSEs.ResultTibble.OMR |> 
  RMSEs.AddRow("User Effect Model", edx.user_effect.RMSE)
```
```{r}
RMSE_kable(RMSEs.ResultTibble.UE)
```

### 2025-12-22 reports/capstone-movielens.site/02.5.UME-model.Rmd

We can now construct predictors and see how much the `RMSE` improves[@IDS2_24-1_ME]:
```{r eval=FALSE}
cv.UM_effect.RMSE <- calc_user_movie_effect_RMSE.cv(cv.UM_effect)

RMSEs.ResultTibble.UME <- RMSEs.ResultTibble.UE |> 
  RMSEs.AddRow("User+Movie Effect Model", cv.UM_effect.RMSE)

@@

  - [calc_user_movie_effect_RMSE.cv](#func.calc_user_movie_effect_RMSE.cv) (described in the *Models Training: Support Functions /* [UME Model: Utility Functions] section) to calculate the *RMSE* for the *Current Model*;

@@

#### UME Model Regularization: Retraining Model with the best $\lambda$
\

@@

Now, we can refine our _User+Movie Effect Model_ by retraining on the entire `edx` dataset with the best value of the $\lambda$ parameter we just figured out (let's call it *Regularized User+Movie Effect Model* or *Regularized UME Model* for short), for the definitive _RMSE_ calculation and use in subsequent models:
```{r, eval=FALSE}
best_result <- UME.rglr.fine_tune.results$best_result
UME.rglr.best_lambda <- best_result["param.best_value"]

>>>>

We can now construct predictors and see how much the `RMSE` improves[@IDS2_24-1_ME].

The following [line of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L513) allows us to perform this task:
```{r eval=FALSE}
cv.UM_effect.RMSE <- calc_user_movie_effect_RMSE.cv(cv.UM_effect)
```

::: {.noteblock data-latex=""}
In the code snippet above, we use the [calc_user_movie_effect_RMSE.cv](#func.calc_user_movie_effect_RMSE.cv) function described in the *Models Training: Support Functions /* [UME Model: Utility Functions] section of [Appendix A](#appndx_a). 
:::

Finally, we add the obtained result to our *Result Table* and print the table using the following [piece of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L):
```{r eval=FALSE}
RMSEs.ResultTibble.UME <- RMSEs.ResultTibble.UE |> 
  RMSEs.AddRow("User+Movie Effect Model", cv.UM_effect.RMSE)

@@

  -  to calculate the *RMSE* for the *Current Model*;

@@

#### UME Model Regularization: Retraining the Model with the best $\lambda$
\

@@

Now, we can refine our _User+Movie Effect Model_ by retraining on the entire `edx` dataset with the best value of the $\lambda$ parameter we just figured out (let's call it *Regularized User+Movie Effect Model* or *Regularized UME Model* for short), for the definitive _RMSE_ calculation and use in subsequent models.

The following [piece of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L664) performs this operation:
```{r, eval=FALSE}
best_result <- UME.rglr.fine_tune.results$best_result
UME.rglr.best_lambda <- best_result["param.best_value"]

put_log1("Re-training Regularized User+Movie Effect Model for the best `lambda`: %1...",
         UME.rglr.best_lambda)

### 2025-12-22 reports/capstone-movielens.site/02.6.UMGE-model.Rmd

#### UMGE Model Regularization: Retraining Model with the best $\lambda$
\

@@

Now, we can refine our _User+Movie+Genre Effect Model_ by retraining on the entire `edx` dataset with the best value of the $\lambda$ parameter we just figured out (let's call it *Regularized User+Movie+Genre Effect Model* or *Regularized UMGE Model* for short), for the definitive _RMSE_ calculation and use in subsequent models:
```{r, eval=FALSE}
  best_result <- UMGE.rglr.fine_tune.results$best_result
  UMGE.rglr.best_lambda <- best_result["param.best_value"]

>>>>

#### UMGE Model Regularization: Retraining the Model with the best $\lambda$
\

@@

Now, we can refine our *User+Movie+Genre Effect (UMGE) Model* by retraining on the entire `edx` dataset with the best value of the $\lambda$ parameter we just figured out (let's call it *Regularized User+Movie+Genre Effect Model* or *Regularized UMGE Model* for short), for the definitive *RMSE* calculation and use in subsequent models.

The following [piece of code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L1076) performs this operation:
```{r, eval=FALSE}
best_result <- UMGE.rglr.fine_tune.results$best_result
UMGE.rglr.best_lambda <- best_result["param.best_value"]

put_log1("Re-training Regularized User+Movie+Genre Effect Model for the best `lambda`: %1...",
         UMGE.rglr.best_lambda)
